Removing nonlinear batch effects using adversarial deep neural networks

### 2019-04-18 supplement.tex
Update the whole paper, particularly the results

\documentclass[notitlepage]{article}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage[square,numbers]{natbib}
\usepackage[nomarkers,nolists]{endfloat}
\renewcommand{\efloatseparator}{\mbox{}} % multiple figures per page
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,footskip=0.25in]{geometry}
\usepackage{csvsimple}
\usepackage{multirow}
\usepackage[table]{xcolor}
\linespread{1.5}
\bibliographystyle{unsrtnat}

\begin{document}

\title{Removing nonlinear batch effects using adversarial deep neural networks}
\author[1]{Jonathan B. Dayton}
\author[1]{Stephen R. Piccolo}
\affil[1]{Department of Biology, Brigham Young University, Provo, UT 84602 USA}
\date{}

\maketitle

\begin{abstract}
	Many biological studies, particularly studies involving the genetics of human disease, could discover items of greater statistical and biological significance if performed with larger data sets.
	Many data sets are publicly available and could be combined in order to increase these studies' significance, but even when the same type of data are collected by multiple studies, minor differences in collection methodologies and environments can lead to confounding effects in the data which can in turn cause incorrect results.
	In order to do so, we applied deep learning to remove these effects and combine molecular biological data sets.
	We created an adversarial variational autoencoder in order to remove confounding effects from the data while minimizing the amount of change to the input data.
	We tested the model on artificially constructed data and commonly used real biological data sets and compared the output with the output from other common batch adjustment algorithms.
	We also applied the model to remove the tissue signal from pan-cancer data.
	Our software is available on GitHub to enable other scientists to remove confounding effects from their own data using our methodologies.
\end{abstract}

\section{Background} \label{sec:background}

When working with biological data, problems can arise from combining data from multiple sources.
For example, if two labs quantify the sequences of the same RNA samples for the same lung cancer patients but use different equipment to do so, each lab will come up with slightly different measurements (see Figure \ref{fig:workflow}).
% Is this ^ the best way to reference figures?
If a third party were to then study the patients using data from both labs, the differences in measurements could alter their analysis.
These systemic effects are known as ``batch effects'' and are understood to have a nontrivial impact on high-throughput biological data \citep{leek_tackling_2010}.
In one study, researchers found that, contrary to previous knowledge, molecular data from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014};
however, referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
Systemic bias can be even more pronounced when using data collected from different tissue types and can skew sec:results, such as in pan-cancer analyses \citep{dayton_classifying_2017-1}.

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow}
	\caption{\textbf{Batch adjustment justification and steps}
	(a) When measurements are collected from a sample ($X$), systematic effects ($H$) also affect the measurements.
	(b) If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
	(c) In order to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
	(d) Second, we remove the estimated effects in order to normalize the batches relative to one another.}
	\label{fig:workflow}
\end{figure}

Several methods exist for removing batch effects from biological datasets.
Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
Since both of these methods use linear adjustments to remove batch effects, they do not account for nonlinear effects.
As machine learning becomes more common in biological research, these nonlinear batch effects become more troublesome since many machine learning algorithms can successfully identify complex interactions between variables.
Advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
These networks have proven effective in removing irrelevant, domain-specific signal in credit rating, online reviews, and image recognition tasks \citep{louizos_variational_2015}.
Recent work \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019} has applied neural networks to batch effects.
However these recent programs have several issues that make them unusable on many real-world datasets:
they require that the input data only contains two batches, that the batches are perfectly balanced, and that the batches are sufficiently large.
These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the documentation for the R SVA package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
Additionally, the statistics these studies have used to validate results don't test whether complex interactions still remain in the data.

To address these issues, we implemented a deep neural network approach to batch correction that works on real-world datasets, and we calculated batch removal measurements based on machine learning algorithms to test for remaining nonlinear effects.
Artificial neural networks are a machine learning tool inspired by the way human brains function; input values pass through layers of linear and nonlinear functions, the final output values are measured against objectives, and the layers of functions are adjusted to bring the outputs closer to the objectives.
This process is repeated until the outputs are sufficiently close to their targets \citep{schmidhuber_deep_2015}.
Autoencoders are a type of neural network that encode and then reconstruct their input, and their traditional objective function is to construct the output as similarly to the input as possible \citep{hinton_reducing_2006}.
Neural networks have historically decreased in effectiveness when working with data from multiple domains \citep{ganin_domain-adversarial_2015}, in part because they may learn based on confounding effects (e.g. which researcher collected the data) instead of learning based on biologically interesting causal effects (e.g. which gene is consistently upregulated in a disease) \citep{louizos_causal_2017-2}.
Recently, researchers have experimented with discouraging the networks from distinguishing between classes by giving them two objective functions:
1. to learn as much as possible about the input data and
2. to forget any patterns that help distinguish between classes \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
This type of dual objective function has been used in conjunction with autoencoders \citep{louizos_variational_2015} where the autoencoder is ``rewarded'' for reconstructing the input faithfully, but ``punished'' for keeping class-specific patterns.
Research has also shown that neural networks are effective in working with biological data; for example, neural networks have been applied to extract biologically relevant latent spaces in RNA-Seq data with as few as 10,000 samples \citep{way_extracting_2017}, and some researchers have used that latent space to generate realistic synthetic biomedical data for other scientific studies \citep{beaulieu-jones_privacy-preserving_2017}.

In this study, we present Confounded, an adversarial autoencoder that identifies and removes effects between different batches.
We test the hypothesis that using an adversarial neural network can correct for batch effects more completely than previous tools do.
We also explore the extent to which batch effects still remain in different datasets after adjustment with various algorithms, and we present a framework to assess how much nonlinear effects between batches still remains after adjustment using various classification algorithms.

\section{Methods} \label{sec:methods}

\subsection{Network Structure}

We used an adversarial autoencoder network to model and remove the batch effects.
This network was structured in two parts: an autoencoder to replicate the input (expression) data and a discriminator to detect remaining batch effects in the autoencoder.
By penalizing the autoencoder for the discriminator's success, the autoencoder subnetwork learned over training to output the expression data with all signs of batch effects removed.
The neural network was implemented in TensorFlow under Python 3.6.
All layers in the network were fully connected and all activation functions were ReLUs \cite{agarap_deep_2018} except the final activations in the autoencoder and the discriminator, which used the sigmoid for their final activations.
We also tried replacing the ReLU activations with SELUs \cite{klambauer_self-normalizing_2017} but saw little if any improvement and therefore returned to using ReLUs.

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/network}
	\caption{\textbf{Network architecture} of Confounded.
	Data with batch effects (represented by different colors) are input into an autoencoder.
	The output of the autoencoder is classified by a discriminator network based on batch.
	The autoencoder is then penalized based on the success of the discriminator.
	Over time, the autoencoder learns to output a faithful representation of the data without signal due to batch.}
	\label{fig:network}
\end{figure}

\subsubsection{Autoencoder}

We initially implemented a basic fully connected autoencoder with two layers for encoding and two layers for decoding and with a smaller (about $\frac{1}{5}$ the size of the input) code layer.
In order to prevent overfitting while allowing sufficient complexity for identifying and removing higher-order batch effects we added dropout to all the layers with a probability of 50\%.
We tested this autoencoder with varying encoding and decoding layer sizes in order to determine the optimal balance of complexity and speed.
We also tested several variations on autoencoders; we implemented a variation on the U-Net \cite{ronneberger_u-net_2015} and plan to test the network with a variational autoencoder \cite{kingma_auto-encoding_2013}.
% TODO: implement and test the variational ae

\subsubsection{Discriminator}

The discriminator is given the output of the autoencoder and was trained to determine the original batch of that output.
It consists of a variable number of fully connected layers of size 500 and a one-hot encoded output for determining batch.
We noticed that the discriminator was overfitting on the data.
Since it memorized which sample belonged to which batch instead of learning batch-specific patterns, the only way for the autoencoder to fool the discriminator was to completely obfuscate the expression data, therefore also removing any relevant signal.
In order to combat this overfitting, we also added 50\%-probability dropout to each layer of the discriminator.
After adding dropout, the network appeared to no longer be overfitting.

\subsubsection{Loss functions}

Three loss functions were used to train the network.
First, the autoencoder was trained on mean squared error \ref{mse} ($MSE_A$) between the input and output layers.
Second, the discriminator was trained on mean squared error ($MSE_D$) between its output and a one-hot encoding of the samples' batch labels.
Finally, the autoencoder was also trained on a combination of the two previous losses,

\begin{equation}
	\label{dual_loss}
	Loss = MSE_A - \lambda{}(MSE_D)
\end{equation}

Where the $\lambda$ value represents a tradeoff parameter for tuning the network's tendency for more faithfully replicating the input or for more completely removing batch effects.

\subsubsection{Training}

In all cases, the network was trained using the Adam Optimizer \cite{kingma_adam_2014} with a training rate of either 0.001 or 0.0001.
We typically trained for 10,000 iterations on mini-batches of size 100.
This training typically took roughly 30 minutes to complete, including time spent to load the input and to save the output.

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/rough/training_loss}
	\caption{Autoencoder and discriminator loss over time for one run of Confounded.
	Over the course of training, the autoencoder more faithfully replicates the input data.
	The autoencoder also seems to introduce noise (see around iteration 3100) in response to the discriminator's slight improvements.}
	\label{fig:training_loss}
\end{figure}

\subsection{Datasets}

In order to test both theoretical and practical differences between Confounded and previous methods, we compared them for a variety of different datasets.
Working with each new dataset enlightened us of a new edge case that hadn't yet been addressed in our code.
By applying to our methods to each of the following datasets, our code became much more robust than if we had only developed and tested on one or two toy datasets.

@@

\subsubsection{MNIST}

The MNIST digits dataset \cite{lecun_mnist_nodate} is a database of images of handwritten digits that are size-normalized and centered.
It contains 60,000 training images and 10,000 test images.
In order to use this dataset, we flattened each 28 by 28 image into a 1D vector of size 784 and put each in a CSV file along with the accompanying digit information.
We limited our dataset to only the 10,000 test images.

By using MNIST, we were able to visually assess how much true signal (in this case, the shape and digit of each handwritten digit image) is preserved after batch adjustment.
Although convolutional layers would typically be used when working with images such as the MNIST dataset, we chose to use only fully connected layers even for this image dataset.
In this way, we show that the autoencoder is still able to find and represent spatial relationships without explicitly defining spatial relationships in the model while testing the same network we use on expression data, where no spatial relationships are inherent.

\paragraph{Synthetic Batch Effects}

Because there is no batch information in the MNIST digits dataset, we had to simulate nonlinear batch effects.
To do so, we wrote a Python script that would take the MNIST data in, apply some nonlinear effect, and output the adjusted data.
We applied nonlinear effects by iteratively realizing vectors of normal values, multiplying and adding these vectors to the ``expression'' vectors, and applying nonlinearity to the adjusted vector.
We split the image data into two batches while keeping the batches balanced (5,000 images for each batch) and including the same number of each digit in each batch.
The same normal vectors were applied to each image in a batch.
Finally, random noise was added to each image in order to prevent images in a batch from being too similar to each other.

\subsubsection{Bladderbatch}

The bladderbatch dataset is a microarray transcriptome expression dataset from a study of patients with bladder cancer \cite{dyrskjot_gene_2004}.
It has been made available in an R package \cite{leek_bladderbatch_2017} and is used in the documentation of the sva R package to illustrate how to batch-adjust using ComBat \cite{leek_sva_2017}.
It contains expression values for 57 samples of patients with and without bladder cancer across 5 unbalanced batches.
Because it is such a small dataset in terms of typical deep learning datasets, we selected it as a way to test whether our network was overfitting.

\subsubsection{GSE37199}

The GSE37199 dataset contains microarray data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
We accessed a version of this data from the Open Science Framework that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
This represents a slightly larger dataset than bladderbatch with only two batches that are closer to being balanced.
Additionally, this dataset has two levels of batches (``centre'' and ``plate'').
We identified this as an item of potential interest: can Confounded adjust for multiple batch effects simultaneously?
However, we left this question for future research and chose instead to focus on the ``plate'' effect for the time being.

\subsubsection{TCGA Pan-cancer Data}

The Cancer Genome Atlas (TCGA) Pan-Cancer project produced expression data for thousands of tumors across many cancer types \cite{the_cancer_genome_atlas_research_network_cancer_2013}.
In a previous study, we used this dataset and attempted to remove the effect of tissue type from this dataset using ComBat \cite{dayton_classifying_2017-1}.
However, we found that a strong nonlinear signal could still be identified in the data after adjustment.
We used the same dataset that we tidied in this study to see if we could remove this nonlinear batch signal more completely.
This dataset has RNA-Seq expression values for 9,365 samples across 25 distinct cancer types.
We used this dataset as a way to test whether Confounded works on RNA-Seq data as well as to test whether we could remove batch effects that we know ComBat cannot.

\subsection{Other adjusters}

We compared our method to two other batch adjusters: a mean and scale adjuster and ComBat \cite{johnson_adjusting_2007}.
We implemented the mean and scale adjuster in the R programming language \cite{r_core_team_r_2014} and used the ComBat implementation from the sva package \cite{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
We wanted to also compare with SVA \cite{leek_capturing_2007} and the Shaham lab's work \cite{shaham_removal_2017,shaham_batch_2018} but were unable to get these two methods to work.
With SVA, the documentation showed how to identify surrogate variables and use them in modeling, but we were unable to find examples of removing the effects that those surrogate variables represent.
We were unable to get the Shaham lab's methods running on our datasets, and we believe that this is due to some underlying assumptions in their code (see \ref{section:mmd} for more information).

\subsection{Statistics and Metrics}

In order to determine how well Confounded removes batch effects as compared to other batch correction software, we calculated several metrics.
First, we calculated mean squared error to see how different the output is from the input.
Second, we calculated sample maximum mean discrepancy to see how differently distributed the batches are from each other.
Third, we determined classification accuracy for several machine learning classifiers unrelated to deep neural networks based on batch in order to determine whether batch can still be identified post-adjustement.
Finally, we determined the same classification accuracies but based on ``true class'' labels to determine how well important signal is maintained after adjustment.

\subsubsection{Mean squared error}

Mean squared error (MSE) is a measure of how much two vectors or matrices deviate from one another.
It is commonly used as a loss value in autoencoders to make the network minimize the difference between the input and output values.
We used MSE as a constraint on the autoencoder in Confounded and wanted to see how well other batch correction software maintain patterns in the input data as measured by MSE.

@@

\subsubsection{Maximum mean discrepancy} \label{section:mmd}

A recent paper also used neural networks to remove batch effects \cite{shaham_removal_2017}.
Instead of constraining the autoencoder to remove batch effects based on a discriminator, these researchers trained their network to minimize maximum mean discrepancy (MMD) between batches in an embedded layer of their network.
We wanted to compare our network against this network, so we calculated MMD for each adjuster.
Unfortunately, we were unable to get their network to work on our datasets for comparison due to some assumptions in their code (there must only be two batches, the two batches must be perfectly balanced, and the dataset must be fairly large).
However, we still calculated this metric to determine whether batches looked like they came from the same distribution after adjustment.

We calculated MMD using the following formula:

@@

\subsubsection{Classification accuracy}

We used four classifiers from the scikit-learn Python library in order to classify on batch and true class before and after adjustment: Naive Bayes, Random Forests, k-Nearest Neighbors, and SVM with a radial basis kernel.
Our true classes were as follows:

\begin{itemize}
	\item MNIST: digit
	\item bladderbatch: presence of cancer
	\item GSE37199: stage of cancer
	\item TCGA: presence of a TP53 mutation
\end{itemize}

We calculated the average of classification accuracies for four-fold cross-validation repeated 10 times.
We interpret lower accuracy for batch classification as meaning that the batch is removed more effectively.
We also interpret higher true class classification as meaning that the important signal is not lost during the process of adjustment.
Therefore, given output data from the ideal batch adjuster, batch classification would be no better than random for any classification algorithm, and true class accuracy would be no lower than accuracy for the unadjusted data.
We report these values in section \ref{sec:results}.

\section{Results} \label{sec:results}

% TODO: put all the figures in here

\subsection{Some batch effects not removed by other adjusters are removed by Confounded}

Various previous studies use differing methods to show qualitatively and quantitatively how well batch effects have been removed.
We used several plots and metrics to compare adjusters, including PCA, t-SNE, MSE, MMD, and classifier batch prediction accuracy (for various classifiers).
Classifier accuracy may be a better indicator of adjuster success than other metrics because many classifiers are able to identify nonlinear interactions;
therefore, a decrease in classifier accuracy represents that nonlinear batch effects have been removed.

PCA and t-SNE plots seem to show a decrease in separability after adjustment with various methods (see Figures \ref{fig:pca} and \ref{fig:tsne})
However, previous research has shown that these plots aren't completely trustworthy in representing nonlinear effects \cite{dayton_classifying_2017-1}.
Additionally, Confounded seems to change the overall shape of the distribution in these two plots, which may indicate an unwanted change to the underlying data.

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/pca}
	\caption{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.}
	\label{fig:pca}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/tsne}
	\caption{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.}
	\label{fig:tsne}
\end{figure}

The MSE and MMD metrics show mixed success (see Figures \ref{mse} and \ref{mmd}).
% TODO: this might not be the case after running with updated code...

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mse}
	\caption{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.}
	\label{fig:mse}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mmd}
	\caption{\textbf{Maximum mean discrepancy} (MMD) between different batches.
	In cases with more than two batches, MMD is computed pairwise between each batch and averaged.}
	\label{fig:mmd}
\end{figure}

Using the batch classification accuracy metric, Confounded seems to outperform other adjusters on larger datasets, whereas ComBat seems to perform better on smaller datasets (see Figure \ref{fig:batch}).
We would expect that after batch adjustment by an ideal adjuster, batch would no longer be detectable.
With both the bladderbatch and GSE37199 datasets, batch classification accuracy decreases well below baseline after batch adjustment with ComBat for all classifiers we tested (see Table \ref{tab:batch}).
Interestingly, batch accuracy also decreases drastically for the MNIST and TCGA datasets, but only for the Naive Bayes classifier.
This may be due to two factors: both ComBat and Naive Bayes use Bayesian methods, so ComBat may specifically remove the effects that Naive Bayes identifies; and Naive Bayes does not find patterns based on interactions between variables.
Although Naive Bayes is no longer able to identify batch effects in the data after ComBat adjustment, Random Forests (which does use interactions between variables) still has a very high accuracy for MNIST and an increased accuracy for TCGA.
In contrast, after adjustment by Confounded, the Random Forests algorithm's accuracy decreases more than with any other adjuster for both the MNIST and TCGA datasets.
This indicates that while ComBat's performance may still be superior for smaller, more traditional expression datasets, Confounded may be better with newer larger datasets.

Although ComBat removes batch well enough to greatly decrease the accuracy of Naive Bayes (a classifier that doesn't account for interactions between variables) in the larger datasets (MNIST and TCGA), Random Forests is still able to predict batch with a high degree of accuracy after ComBat adjustment.

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy}
	\caption{\textbf{Batch classification accuracy} Accuracies for batch classification for 4-fold cross validation repeated 10 times for several classifiers.
	See also Table \ref{tab:batch}.}
	\label{fig:batch}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/batch.tex}
	\caption{\textbf{Batch classification accuracy} for several datasets and adjusters.
	The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
	See also Figure \ref{fig:batch}.}
	\label{tab:batch}
\end{table}

Confounded greatly decreases batch accuracy with the TCGA dataset.

Additional modifications to the network may increase Confounded's efficacy with the larger datasets and possibly with the smaller ones as well.

\subsection{Important signal is still detectable after adjustment by Confounded}

With the smaller datasets, Confounded seems to keep true class information roughly as well as ComBat.
Figure \ref{fig:true_class}, Table \ref{tab:true_class}

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy}
	\caption{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross validation repeated 10 times.
	See also Table \ref{tab:true_class}.}
	\label{fig:true_class}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/true_class.tex}
	\caption{\textbf{True class classification accuracy} for several datasets and adjusters.
	After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
	See also Figure \ref{fig:true_class}.}
	\label{tab:true_class}
\end{table}

With the larger datasets, Confounded seems to be losing some accuracy.
Figure \ref{fig:mnist}

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/rough/mnist}
	\caption{\textbf{MNIST handwritten digits} (a) before any adjustment,
	(b) with artificial noise added,
	(c) adjusted for noise by ComBat, and
	(d) adjusted for noise by Confounded.
	Although Confounded seems to remove more noise from the background and make it look more like the original background, it struggles in some cases to accurately replicate the input data.}
	\label{fig:mnist}
\end{figure}

Loss of accuracy may be ameliorated by longer training.
Or it might be an acceptable amount of lost accuracy, depending on the application.

\section{Discussion} \label{sec:discussion}

Most exciting takeaways

\begin{enumerate}
	\item Nonlinear interactions still remain after ComBat for larger datasets
	\begin{enumerate}
		\item Random Forests often has high batch classification accuracy after ComBat-adjustment.
	\end{enumerate}
	\item Simple metrics aren't enough to identify the nonlinear interactions, but different classifiers help
	\begin{enumerate}
		\item Figs \ref{fig:mmd} and \ref{fig:mse} both show that ComBat performed better than Confounded for the MNIST dataset, but the effects left after ComBat finished are still identifiable by Random Forests with 99.9\% accuracy (Table \ref{tab:batch}).
	\end{enumerate}
	\item ComBat works really well for small, traditional datasets % include flip side for #s 3&4
	\item Confounded removes nonlinear effects in larger datasets
	\item Confounded shows promise for other confounders, not just batch effects (could work with combining rnaseq + microarray) and seems to work with multiple types of data
\end{enumerate}

\bibliography{references}

\end{document}

>>>>

We also tried replacing the ReLU activations with SELUs \cite{klambauer_self-normalizing_2017} but saw little if any improvement and therefore returned to using ReLUs.

@@

Autoencoder

In order to prevent overfitting while allowing sufficient complexity for identifying and removing higher-order batch effects we added dropout to all the layers with a probability of 50\%.
We tested this autoencoder with varying encoding and decoding layer sizes in order to determine the optimal balance of complexity and speed.
We also tested several variations on autoencoders; we implemented a variation on the U-Net \cite{ronneberger_u-net_2015} and plan to test the network with a variational autoencoder \cite{kingma_auto-encoding_2013}.

Discriminator

We noticed that the discriminator was overfitting on the data.
Since it memorized which sample belonged to which batch instead of learning batch-specific patterns, the only way for the autoencoder to fool the discriminator was to completely obfuscate the expression data, therefore also removing any relevant signal.

Datasets

Working with each new dataset enlightened us of a new edge case that hadn't yet been addressed in our code.
By applying to our methods to each of the following datasets, our code became much more robust than if we had only developed and tested on one or two toy datasets.

@@

GSE37199

Additionally, this dataset has two levels of batches (``centre'' and ``plate'').
We identified this as an item of potential interest: can Confounded adjust for multiple batch effects simultaneously?
However, we left this question for future research and chose instead to focus on the ``plate'' effect for the time being.

Other methods

We wanted to also compare with SVA \cite{leek_capturing_2007} and the Shaham lab's work \cite{shaham_removal_2017,shaham_batch_2018} but were unable to get these two methods to work.
With SVA, the documentation showed how to identify surrogate variables and use them in modeling, but we were unable to find examples of removing the effects that those surrogate variables represent.
We were unable to get the Shaham lab's methods running on our datasets, and we believe that this is due to some underlying assumptions in their code (see \ref{section:mmd} for more information).

MSE

@@

MMD

We wanted to compare our network against this network, so we calculated MMD for each adjuster.
Unfortunately, we were unable to get their network to work on our datasets for comparison due to some assumptions in their code (there must only be two batches, the two batches must be perfectly balanced, and the dataset must be fairly large).

following formula:

@@
